Diversity and evolution of centromere repeats in the maize genome

### 2014-07-16 centc.tex
PB final manuscript for JRI to look over changes

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  
Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}, with centromeres 1, 2, 5, 7, 9, and 10 deriving from subgenome 1 and 3, 4, 6, 8 from subgenome 2.  
Due to the maize iMap resource \citep{zhou2009single}, we believe large clusters are less likely to be misplaced within the genome.
Therefore, we focused our analyses on the 52 clusters $>$10 Kb in length (Supplementary Fig \ref{Supp_clusterhist}).
We assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  
Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable).
If we restrict the analysis to clusters  $>$20 Kb with clear assignment to one subgenome, all 16 clusters were found in subgenome 1.  
Even correcting for the genome-wide overrepresentation of subgenome 1 (64.7\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for both 10 Kb and 20 Kb clusters).
One cluster $>$20 Kb falls within an unassigned region on chromosome 3.
This difference between the subgenomes is robust to different criteria of cluster size and distance (Table \ref{tab:suppvartandem}).

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
\includegraphics{Fig1_circos}
\caption{CentC repeat location in relation to the maize subgenomes.  The outer ring depicts chromosomal assignment to the two subgenomes, with higher confidence regions in darker colors.  Green corresponds to subgenome 1, brown to subgenome 2.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20 Kb in length}
\label{circos}    
\end{figure}

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\centering
\includegraphics[width=1\textwidth]{Fig4_kelly2}
\caption{(a) FISH analysis of a single individual heterozygous for B73 and \emph{Zea luxurians} (GRIN accession PI422162).
Chromosomes, ordered from 1 (left) to 10 (right), were hybridized with the \citet{kato2004paint} probe cocktail. 
Green shows CentC and a 4-12-1 subtelomere repeat, blue the 180 bp knob repeat, red the abundant TAG microsatellite and another subtelomeric repeat, white the TR1 knob repeat, orange the Cent4 repeat, yellow the 5S rDNA repeat, and aqua shows the Nucleolus Organizer Region.
The green signals at the primary constrictions (arrows) are CentC.
Note that \emph{Z. luxurians} has far brighter CentC signals.
This image was graciously provided by Patrice Albert.
(b) FISH analysis of a single individual heterozygous for B73 and \emph{Z. parviglumis}(GRIN accession PI566687).
Chromosomes were hybridized with the \citet{Shi2010} probe cocktail showing CentC in red and CRM2 in green. 
Each separate probe is shown separately below the two-color image, highlighting that CentC is more abundant in \emph{Z. parviglumis} and CRM2 is more abundant in maize. 
(c)  Mb of CentC in genomic libraries of maize and teosinte. 
Box plots show data from \citet{Chia2012}. 
Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}. For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot.
}
\label{abundance}    
\end{figure}

@@

We find that virtually all the large arrays of CentC in the maize  genome derive from one of the two ancestral genomes present in modern day \emph{Zea} (Figure \ref{circos}, Table \ref{tab:suppvartandem}).  
This biased ancestry mirrors differences in genic expression and deletion seen between the subgenomes  \citep{Schnable2011}. Higher deletion rates on subgenome 2 may explain our observation, but the finding of small regulatory RNA’s corresponding to centromeric repeats \citet{ReinhartBartel2002} in other taxa may suggest a more active mechanism behind the observed differences.

@@

In conclusion, our detailed study of centromere repeats in the B73 maize genome has highlighted differential contribution of subgenome, spatial autocorrelation along a chromosomes, and changes in abundance over the short time scale of domestication.  
Since maize lines show vast cytogentic variation, further work evaluating CentC evolution across multiple populations and multiple related species may shed additional light on the timing and causes of these changes.

>>>>

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  
Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}, with centromeres 1, 2, 5, 7, 9, and 10 deriving from subgenome 1 and 3, 4, 6, 8 from subgenome 2.  
Due to the maize iMap resource \citep{zhou2009single}, large clusters should be less likely to be misplaced within the genome.
Therefore, we focused our analyses on the 52 clusters $>$10 Kb in length (Supplementary Fig \ref{Supp_clusterhist}).
We assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  
Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable).
If we restrict the analysis to clusters  $>$20 Kb with clear assignment to one subgenome, all 16 clusters were found in subgenome 1.  
Even correcting for the genome-wide overrepresentation of subgenome 1 (64.7\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for total CentC, 10 Kb, and 20 Kb clusters).
One cluster $>$20 Kb falls within an unassigned region on chromosome 3.
This difference between the subgenomes is robust to different criteria of cluster size and distance (Table \ref{tab:suppvartandem}).

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
\includegraphics[width=10cm,height=10cm,keepaspectratio]{Fig1_circosv3.png}
\caption{CentC repeat location in relation to the maize subgenomes.  The outer ring depicts chromosomal assignment to the two subgenomes, with higher confidence regions in darker colors.  Green corresponds to subgenome 1, brown to subgenome 2.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20 Kb in length.  Asterisks indicate approximate location of the centromere}
\label{circos}    
\end{figure}

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{Fig4_kelly2}
\caption{(a) FISH analysis of a single individual heterozygous for B73 and \emph{Zea luxurians} (GRIN accession PI422162).
Chromosomes, ordered from 1 (left) to 10 (right), were hybridized with the \citet{kato2004paint} probe cocktail. 
Green shows CentC and a 4-12-1 subtelomere repeat, blue the 180 bp knob repeat, red the abundant TAG microsatellite and another subtelomeric repeat, white the TR1 knob repeat, orange the Cent4 repeat, yellow the 5S rDNA repeat, and aqua shows the Nucleolus Organizer Region.
The green signals at the primary constrictions (arrows) are CentC.
Note that \emph{Z. luxurians} has far brighter CentC signals.
This image was graciously provided by Patrice Albert.
(b) FISH analysis of a single individual heterozygous for B73 and \emph{Z. parviglumis}(GRIN accession PI566687).
Chromosomes were hybridized with the \citet{Shi2010} probe cocktail showing CentC in red and CRM2 in green. 
Each separate probe is shown separately below the two-color image, highlighting that CentC is more abundant in \emph{Z. parviglumis} and CRM2 is more abundant in maize. 
(c)  Mb of CentC in genomic libraries of maize and teosinte. 
Box plots show data from \citet{Chia2012}. 
Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}. For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot.
}
\label{abundance}    
\end{figure}

@@

We find that virtually all the large arrays of CentC in the maize reference genome derive from one of the two ancestral genomes present in modern day \emph{Zea} (Figure \ref{circos}, Table \ref{tab:suppvartandem}).  
This biased ancestry mirrors differences in genic expression and deletion seen between the subgenomes  \citep{Schnable2011}. 
Higher deletion rates on subgenome 2 may explain our observation, but the finding of small regulatory RNA’s corresponding to centromeric repeats \citet{ReinhartBartel2002} in other taxa may suggest a more active mechanism behind the observed differences.

@@

In conclusion, our detailed study of centromere repeats in the B73 maize genome has highlighted differential contribution of subgenome, spatial autocorrelation along a chromosomes, and changes in abundance over the short time scale of domestication.  
Since maize lines show vast cytogentic variation, further work evaluating CentC evolution across multiple populations and multiple taxa may shed additional light on the timing and causes of these changes.
